flowCore: A Bioconductor software package for high throughput flow cytometry analysis

### 2008-08-12 paper/manuscript.tex
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33197 bc3139a8-67e5-0310-9ffc-ced21a209358

\title{Outline flowCore paper}

\author{Florian Hahne*\\
  Nolwenn LeMeur*\\
  Byron Ellis\\
  Ryan Brinkman\\
  Perry Haaland\\
  Errol Strain\\
  Deepayan Sarkar\\
  Josef Spidlen\\
  Robert Gentleman
 }

@@

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
During the last several years, automation technologies have been
developed that enable the use of flow cytometry high content screening
(FC-HCS) to generate large, complex datasets in both basic and
clinical research applications \citep{brinkman2007hcf}. A serious
bottleneck in the interpretation of existing studies and the
application of FC-HCS to even larger, more complex problems is that
data management and data analysis methods have not advanced
sufficiently far from the methods developed for applications of flow
cytometry (FCM) to small-scale, tube-based studies.

@@

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In particular, the Bioconductor project
provides R software modules for biological and clinical data analysis.
It is widely used to process such data, spanning a diversity of
research fields including genomics, proteomics and cell biology
\citep{BIOC}. The R language and Bioconductor are
particularly geared towards the analysis of large complex datasets
that typically arise from high throughput experiments such as
microarray gene expression analysis or FC-HCS. Using flow cytometry as
a high-throughput technology requires suitable software infrastructure
to facilitate the rapid handling of the data. In our implementation of
\Rpackage{flowCore} we rely on two important lessons learned from the field of
gene expression data analysis: the first being the importance of data
structures that reflect the underlying data and facilitate the
manipulations that are of most interest, while the second is the
importance of a modular architecture that allows for many developers
to extend and use the underlying infrastructure and to combine tools
in complex work flows.

@@

In \Rpackage{flowCore}, gating operations are represented by classes
that can be extended in an object-oriented manner
(Table~\ref{table2}). Basic gate types such as rectangular gates,
ellipses and polygon gates are implemented as part of the
framework. In addition, we introduce the notion of data-driven gates,
or filters, for which the necessary parameters are computed based on
the properties of the underlying data, for instance by modelling data
distribution or by density estimation. This approach is fundamentally
different from the traditional application of static gating regions
across samples, as it is able to take into accounts unforeseen changes
in signal intensities. \textit{e.g.} drifts in the instrumentation
over time or sample variablity.

@@

Data analysis for most experiments usually begins with a quality
assurance step, and we can use functionality from the \Rpackage{flowQ}
package to create an HTML report that highlights potential quality
issues. Assuming that the data has already been imported as the
\Rclass{flowSet} object ``dat'', the following simple line of code
produces the quite complex output:

@@

The report is interactive and provides drill-down to more detailed
aspects of the analysis, starting from a concise overview
(Figure~\ref{flowQ}). The design of \Rpackage{flowCore}'s data model
allows for a coherent treatment of all the samples, hence we are able
to compare features between individuals, or between groups of
individuals, based on the available metadata information.

@@

Through \Rpackage{flowCore}, we have provided the FCM community a
freely available, highly functional, standards compliant development
and analysis platform for high throughput data analysis.  Our
intention is to provide a platform for the collaborative development
of new analysis methods that will facilitate the transition of these
new methods to the larger flow community.  Our experience has been
that the collaborative effort we enable for devising new methodology
has been proven beneficial for a number of different biological and
computational biology challenges.  We hope that our framework will be
the foundation for fruitful shared research by many collaborators from
multiple scientific fields and will help resolve bottlenecks that
currently prevent the further development of and deployment of FC-HCS
to increasingly complex and important scientific and clinical
applications.

\section*{Acknowledgements}
This work was supported by NIH grant EB005034. RRB is a Michael Smith
Foundation for Health Research Scholar and a International Society for
Analytical Cytology Scholar.

>>>>

\title{flowCore: A Bioconductor software package for high throughput flow cytometry analysis}

\author{Florian Hahne*\\
  Nolwenn LeMeur*\\
  Byron Ellis\\
  Ryan R. Brinkman\\
  Perry Haaland\\
  Errol Strain\\
  Deepayan Sarkar\\
  Josef Spidlen\\
  Robert Gentleman
 }

@@

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
During the last several years, automation technologies have been
developed that enable the use of FCM high content screening
(FC-HCS) to generate large, complex datasets in both basic and
clinical research applications \citep{brinkman2007hcf}. A serious
bottleneck in the interpretation of existing studies and the
application of FC-HCS to even larger, more complex problems is that
data management and data analysis methods have not advanced
sufficiently far from the methods developed for applications of flow
cytometry (FCM) to small-scale, tube-based studies.

@@

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In particular, the Bioconductor project
provides R software modules for biological and clinical data analysis.
It is widely used to process such data, spanning a diversity of
research fields including genomics, proteomics and cell biology
\citep{BIOC}. The R language and Bioconductor are
particularly geared towards the analysis of large complex datasets
that typically arise from high throughput experiments such as
microarray gene expression analysis or FC-HCS. Using FCM as
a high-throughput technology requires suitable software infrastructure
to facilitate the rapid handling of the data. In our implementation of
\Rpackage{flowCore} we rely on two important lessons learned from the field of
gene expression data analysis: the first being the importance of data
structures that reflect the underlying data and facilitate the
manipulations that are of most interest, while the second is the
importance of a modular architecture that allows for many developers
to extend and use the underlying infrastructure and to combine tools
in complex work flows.

@@

In \Rpackage{flowCore}, gating operations are represented by classes
that can be extended in an object-oriented manner
(Table~\ref{table2}). Basic gate types such as rectangular gates,
ellipses and polygon gates are implemented as part of the
framework. In addition, we introduce the notion of data-driven gates,
or filters, for which the necessary parameters are computed based on
the properties of the underlying data, for instance by modelling data
distribution or by density estimation. This approach is fundamentally
different from the traditional application of static gating regions
across samples, as it is able to take into accounts unforeseen changes
in signal intensities (\textit{e.g.,} drifts in the instrumentation
over time or sample variablity).

@@

Data analysis for most experiments usually begins with a quality
assurance step, and we can use functionality from the \Rpackage{flowQ}
package to create an HTML report that highlights potential quality
issues. Assuming that the data has already been imported as the
\Rclass{flowSet} object ``dat'', the following simple line of code
produces the output shown in Figure~\ref{flowQ}:

@@

The report is interactive and provides drill-down to more detailed
aspects of the analysis, starting from a concise overview. The design of \Rpackage{flowCore}'s data model
allows for a coherent treatment of all the samples, hence we are able
to compare features between individuals, or between groups of
individuals, based on the available metadata information.

@@

Through \Rpackage{flowCore}, we have provided the FCM community an open source,
freely available, highly functional, standards compliant development
and analysis platform for high throughput data analysis.  Our
intention is to provide a platform for the collaborative development
of new analysis methods that will facilitate the transition of these
new methods to the larger flow community.  Our experience has been
that the collaborative effort we enable for devising new methodology
has been proven beneficial for a number of different biological and
computational biology challenges.  We hope that our framework will be
the foundation for fruitful shared research by many collaborators from
multiple scientific fields and will help resolve bottlenecks that
currently prevent the further development of and deployment of FC-HCS
to increasingly complex and important scientific and clinical
applications.

\section*{Acknowledgements}
This work was supported by NIH grant EB005034 and by the Michael Smith
Foundation for Health Research. RRB is an ISAC Scholar.
